---
title: 'Intro to R: The Review!'
author: "[your name here]"
date: "18 January 2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

We learned _a lot_ of skills last quarter in DAnC. You should be proud of yourselves! This is a little review of everything we learned.  

# 1. Assignment operator, functions, and subsetting

We'll start with the basics: assigning values to objects using the assignment operator `<-`. 

```{r}

apple <- 5

orange <- 3

```

Then, we learned how to use functions in R. There are many functions out there to do the work you want!

```{r}

sum(apple, orange)

```

However, we're not stuck with simple numbers! We can get more complex with our objects.

```{r}

apple_weights <- c(0.3, 0.2, 0.25, 0.29)

orange_weights <- c(0.2, 0.18, 0.25, 0.24)

fruits <- c("apple", "orange", "banana", "pear", "persimmon", "grapefruit")

```

Then, we can subset our data using `[]`. Let's say we want the first apple weight.

```{r}

apple_weights[1]

```

Now, let's say we want all the orange weights that are less than 0.24.
```{r}

orange_weights[orange_weights < 0.24]

```

What would you write if you wanted to get the third element in `fruits`?

```{r}



```

# 2. More complex data

We're going to do a bit of set up by loading in our packages (aka libraries) and our data. Remember that packages are collections of functions that do specific things. Today, we're using the `tidyverse`, which is a suite of packages all designed to make your data analysis life easier.  
For something fun this time (not that urchins aren't fun, we just worked with them for 3 months), we're going to be using data from Tidy Tuesday, a weekly social coding event where people work to visualize data from a dataset and share on Twitter! This data was from June 2020 on marble races; check out the [readme](https://github.com/rfordatascience/tidytuesday/blob/master/data/2020/2020-06-02/readme.md) for info on the columns and what kinds of data they hold!

```{r set up}

##################### libraries ##################### 

# if you don't have the tidyverse yet, uncomment the code below and run it first!
# install.packages("tidyverse")
library(tidyverse)

####################### data ######################## 

marbles <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2020/2020-06-02/marbles.csv')

```

# 3. Filtering observations from a data frame and the `%>%` operator 

We have our `marbles` data loaded into RStudio. Let's say we want to know the mean race time for a specific marble - maybe the one called "Snowflake."  

We can do this using functions and operators from the `tidyverse`, which is the package we loaded in. 

The way we can get the mean race time for Snowflake with `tidyverse` functions is:  
1. `filter()`: filter the data frame to only include observations for Snowflake  
2. `summarize()`: calculate the mean race time  

Both of these functions are from the `dplyr` package in the `tidyverse`. 

We also learned how to use the pipe operator `%>%` from the `magrittr` package. This allows you to link functions together, which can be very handy! This is a pretty simple example, but we learned how powerful that operator can be (and will demonstrate later on).  

```{r snowflake}

snowflake_mean <- marbles %>% 
  
  filter(marble_name == "Snowflake") %>% 
  
  summarize(mean_time = mean(time_s))


snowflake_mean
```

Try finding the mean time for another marble in the data set! 

```{r your-turn}






```

We could do this for every marble in the data frame, but that would get tedious. To streamline things, we can use the `group_by()` function also from `dplyr` to tell R: think of this as a _group_.  

```{r using-group_by()}

all_means <- marbles %>% 
  
  group_by(marble_name) %>% 
  
  summarize(mean_time = mean(time_s))

```

# 4. Data visualization

Now that we have our data, it's time for the fun part: visualizing it!  

Let's say we're interested in the way that average time per lap is related to track length.  

This is the most basic way to visualize it:

```{r basic-visualization}

ggplot(data = marbles, aes(x = track_length_m, y = avg_time_lap)) +
  
  geom_point() 

```

But... we can make this a lot more readable!  

```{r readable-visualization}

ggplot(data = marbles, aes(x = track_length_m, y = avg_time_lap)) +
  
  geom_point(size = 2) +
  
  geom_smooth(color = "royalblue", size = 2) +
  
  theme_bw() +
  
  labs(x = "Track length (m)",
       y = "Average time per lap (s)",
       title = "How does track length influence average time per lap?",
       caption = "Data courtesy of Thomas Mock, #TidyTuesday",
       color = "Team name") 

```

Ok, that was a pretty good plot. But let's say we're interested in the best marble. 

We'll filter out observations to only include races with 1 lap. Then, we'll calculate the mean race time for each marble, and do some arrangement so that we can see the order better.

```{r best-marble}

best_marble <- marbles %>% 
  
  filter(number_laps == 1) %>% 
  
  group_by(marble_name) %>% 
  
  summarize(mean_times = mean(time_s)) %>% 
  
  arrange(-mean_times) %>% 
  mutate(marble_name = fct_relevel(marble_name, marble_name))
```

Then, we can visualize the data! Again, you could do something basic:

```{r basic-column}

ggplot(data = best_marble, aes(x = marble_name, y = mean_times)) +
  
  geom_col()

```

Or, you could do something fun to look at!  

First, we'll load in `viridis`, which is a package of color palettes to make choosing colors easier. There are **tons** of color palette packages out there, so lots to explore!

```{r load-viridis}

# if you don't have viridis yet, uncomment the line below and run it first!
# install.packages("viridis")
library(viridis)

```

Then, we can plot!

```{r prettier-column}

ggplot(data = best_marble, aes(x = marble_name, y = mean_times)) +
  
  geom_col(aes(fill = marble_name)) +
  
  scale_y_continuous(expand = c(0, 0), limits = c(0, 35)) +
  
  scale_fill_viridis(option = "viridis", discrete = TRUE) +
  
  labs(x = "Marble name", 
       y = "Average time per lap (s)",
       title = "Which marble is the fastest?",
       subtitle = "Prim is the winner!",
       caption = "Data courtesy of Thomas Mock, #TidyTuesday") + 
  
  theme_minimal() +
  
  theme(legend.position = "none", 
        axis.text.x = element_text(angle = 60, vjust = 0.6))

```


