---
title: "UCSB DAnC Intro to R part 3"
author: "[your name here]"
date: "3 December 2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# 0. Set up
```{r}
# libraries
library(tidyverse)
library(janitor)

# data
urchins <- read_csv(here::here("2020-12-03-intro_to_R-part03", "sea-urchin.csv"))
```

# 1. Review of Intro to R part 2: 

## a. cleaning up column names

Here, we cleaned up those column names so that they were easier to type out - a small step, but important down the line.

```{r}
# use janitor::clean_names() function to change the column names to 
# 1) not have spaces and 2) not be capitalized
urchins2 <- urchins %>% 
  clean_names() 

urchins2
```

## b. Basic statistics

Now, we can calculate the mean mass of urchins - again, something you can do in Excel, but is **reproducible** when you're doing it in R. 

```{r}
# calculate mean mass of urchins using dplyr::summarize() function
av_mass <- urchins2 %>% 
  summarize(mean = mean(sea_urchin_mass_g))  

av_mass
```

## c. Using `dplyr::filter()` function to subset data

Let's say we only want to calculate the mean mass for urchins in the species that we're actually interested in. We'll start with _Echinometra mathaei_.
```{r}
# filter urchins to only include the species of interest, "Echinometra mathaei"
EM_only <- urchins2 %>% 
  # filter for the species of interest
  filter(sea_urchin_species == "Echinometra mathaei")

# look at the data frame
view(EM_only)

# calculate mean mass
EM_mass <- EM_only %>% 
  summarize(mean = mean(sea_urchin_mass_g))

# look at the main mass of E. mathaei
view(EM_mass)
```

Remember that you can use the pipe operator to string those functions together.

```{r}
# so what you're telling R is to:
# 1. use the urchins2 data frame
EM_mass2 <- urchins2 %>% 
  # 2. filter the data frame to only include E. mathaei
  filter(sea_urchin_species == "Echinometra mathaei") %>%
  # 3. calculate the mean mass
  summarize(mean = mean(sea_urchin_mass_g))

# print the output of that pipe
EM_mass2
```

Now, we can do the same thing with the other urchin species, _Diadema savignyi_.
```{r}
# doing the same thing with the other urchin species, Diadema savignyi.
DS_mass <- urchins2 %>% 
  filter(sea_urchin_species == "Diadema savignyi") %>% 
  summarize(mean = mean(sea_urchin_mass_g))

DS_mass
```

## d. Creating the final data frame

We're using the word "final" to refer to the data frame that you want to get to in order to start making figures from the data.

```{r}
# 1. use urchins2 data frame
final <- urchins2 %>% 
  # use group_by() to tell R to consider different groups in the data:
  # 1) reef habitat, and 2) urchin species
  group_by(reef_habitat, sea_urchin_species) %>% 
  # use summarize() to calculate the mean urchin mass and the standard error
  summarize(mean = mean(sea_urchin_mass_g), 
            err = sd(sea_urchin_mass_g)/sqrt(length(sea_urchin_mass_g))) 

final
```

# 2. Data visualization

To create a bar graph, we'll use the `ggplot2` package in the Tidyverse. You can make graphs in baseR (the built-in option), but `ggplot2` allows a lot more customization.

```{r}
# main function: ggplot()
# ggplot() takes a lot of arguments
# main arguments: data frame, aes()
# aes() stands for aesthetics

# telling ggplot() to use `final` as the data
mass_plot <- ggplot(data = final, 
                    # telling ggplot to use reef_habitat as x-axis
                    aes(x = reef_habitat, 
                        # use mean as the y-axis
                        y = mean, 
                        # color by sea urchin species
                        fill = sea_urchin_species)) +
  
  # geom_col: geometry that is a column graph
  # position = "dodge": put columns next to each other instead of on top of each other
  geom_col(position = "dodge", width = 0.8) +
  
  # geom_errorbar: adds error bars to your column graph
  # first, tell geom_errorbar where error bars start and end
  geom_errorbar(aes(ymin = mean - err, ymax = mean + err), 
                position = position_dodge(0.8), 
                width = 0.3)

mass_plot
```

In this example, the `fill` call tells R, "These are groups that I want to graph separately."

We won't go through all the code for this graph, but you can format this professionally for publication using functions within `ggplot()`.

```{r}
# takes same aes() as previous plot
mass_plot_pub <- ggplot(final, aes(x = reef_habitat, y = mean, fill = sea_urchin_species)) +
  
  # still making a column graph
  geom_col(position = "dodge", width = 0.8, color = "black") +
  
  # still plotting standard error
  geom_errorbar(aes(ymin = mean - err, ymax = mean + err), position = position_dodge(0.8), width = 0.3) +
  
  # allows you to customize the y-axis
  scale_y_continuous(expand = c(0,0), limits = c(0, 24)) +
  
  # allows you to customize the colors that go into the fill
  scale_fill_manual(values = c("#616161", "#adadad")) +
  
  # built in themes
  # others: theme_bw(), theme_dark()
  # theme(): allows customization of your own theme
  theme_minimal() + 
  
  # change labels
  labs(x = "Reef Habitat",
       y = "Average mass (g)", 
       fill = "Urchin species")

mass_plot_pub
```

We can also create a scatter plot. The first thing we'll have to do is filter the `urchins2` data frame to only include observations from the back reef.

```{r}
back_reef <- urchins2 %>% 
  # filter to only include urchins from back reef
  filter(reef_habitat == "Back Reef")

back_reef
```

Now we can use this data frame to feed into ggplot.

```{r}
scatter <- ggplot(back_reef, 
                  aes(x = sea_urchin_mass_g, y = sea_urchin_spine_length_cm)) +
  
  # geom_point(): create a scatter geometry
  geom_point() +
  
  # geom_smooth(): creates a trend line geometry
  geom_smooth(method = lm)

scatter
```

Here's the code for making this plot look professional:

```{r}
scatter_pub <- ggplot(back_reef, aes(x = sea_urchin_mass_g, y = sea_urchin_spine_length_cm)) +
  
  # change the size of the points to be bigger
  geom_point(size = 2) +
  
  # change the color line to be black
  geom_smooth(method = lm, colour = "black") +
  
  # take away the grey background and make gridlines grey
  theme_minimal() +
  
  # change labels
  labs(x = "Sea urchin mass (g)",
       y = "Sea urchin spine length (cm)")

scatter_pub
```

# 3. Cooler visualizations

What if you wanted to see what trends between mass and spine length were for both species in the back reef?

```{r}
scatter_new <- ggplot(urchins2, aes(x = sea_urchin_mass_g, y = sea_urchin_spine_length_cm, group = sea_urchin_species)) +
  
  # facet_wrap allows separation of plots
  facet_wrap(~reef_habitat) +
  
  # create a scatter plot
  # alpha: changes transparency
  geom_point(aes(shape = sea_urchin_species), size = 2, alpha = 0.8) +
  
  # customize shapes from geom_point
  scale_shape_manual(values = c(16, 2)) +
  
  # add trendlines, change the color to be red
  geom_smooth(method = lm, color = "red") +
  
  # get rid of grey background, make gridlines grey
  theme_minimal() +
  
  # clean up plot labels
  labs(x = "Sea urchin mass (g)", 
       y = "Sea urchin spine length (cm)",
       shape = "Sea urchin species")

scatter_new
```

RStudio has published many cheat sheets for commonly used packages. The one for `ggplot2` outlines a lot of the functions that you'll need to make professional looking figures: https://rstudio.com/wp-content/uploads/2015/03/ggplot2-cheatsheet.pdf

And if you're interested in experimenting with color, there is a reference sheet for that too: http://sape.inf.usi.ch/quick-reference/ggplot2/colour

Other useful functions in dplyr
```{r}
# select
urchins3 <- select(urchins2, reef_habitat, sea_urchin_species, sea_urchin_mass_g)
urchins3

# mutate
urchins4 <- urchins2 %>%
  mutate(standardized_mass_g = (sea_urchin_mass_g - mean(sea_urchin_mass_g)) / sd(sea_urchin_mass_g))  %>%
  mutate(standardized_spine_cm = (sea_urchin_spine_length_cm - mean(sea_urchin_spine_length_cm) / sd(sea_urchin_spine_length_cm) )) 

urchins4

scatter_new2 <- ggplot(urchins4, aes(x = standardized_mass_g, y = standardized_spine_cm, group = sea_urchin_species)) +
  facet_wrap(~reef_habitat) +
  geom_point(aes(shape = sea_urchin_species, color = sea_urchin_species), size = 2, alpha = 0.8) +
  scale_shape_manual(values = c(21, 22)) +
  geom_smooth(method = lm, color = "red") +
  theme_minimal() +
  labs(x = "Sea urchin mass (g)", 
       y = "Sea urchin spine length (cm)",
       shape = "Sea urchin species", 
       color = "Sea urchin species") 

scatter_new2


scatter_new
```