---
title: "Introduction to R - DANC"
author: "[your name here]"
date: "19 March 2020"
output: html_document
---

Logistics 
1. No meeting next week (Thanksgiving)
2. How does this time work for everyone?
3. Student leadership positions and community building 
4. Announcement from Jordan about mentoring opportunities

You'll notice that we are in a .Rmd file. Last week we used a .R file. 

What's the difference? 

```{r setup, include=FALSE}
# include: Whether to include anything from a code chunk in the output document. When include = FALSE, this whole code chunk is excluded in the output, but note that it will still be evaluated if eval = TRUE. 
knitr::opts_chunk$set(echo = TRUE)
# echo: Whether to echo the source code in the output document (someone may not prefer reading your smart source code but only results)
```

This is an RMarkdown file. It's a very useful format to use if you want to run code in between paragraphs of writing. **Whole scientific papers can be written in RMarkdown**, so keep this in mind for any future work!

We'll start sectioning off our document now.

Let's warm up our brains and do a quick review from two weeks ago: 

```{r}
#######################
# Naviagting R Studio #
#######################
# Top Left : where we write and can save our code (where we keep our recipes)
# Bottom Left : where we execute our code. important = code written here will not be saved. (where we cook our food)
# Top Right : where our data and model outputs are (kitchen counter)
# Bottom right : files that we can specify in our code (cupboard)

# Recall what the pound sign does 
# you can write comments in the R Script by putting a pound sign at the beginning of the line
print(c("this", "is", "code"))
# this is not code 

#######################
# Variable assignment #
#######################
# We'll start with some basics. We'll assign values to objects.
# assign the number 5 to an object called spine_length
spine_length <- 5

#############
# Functions #
#############
# Functions allow us to run specific tasks in R 
# Recall  that packages have functions that aren't already built into R (we'll cover these today!)
# We can start by calculating the square root of spine_length.
root <- sqrt(spine_length)

# Reminder: How might you find more about the sqrt() function if you didn't know the arguments that could go inside the parentheses? 
?sqrt
help("sqrt")

########
# Data #
########
# We've assigned values to objects, but this is especially powerful when working with data. We can assign anything to an object. 
# We'll create a vector of names of marine animals in an object called marine_animals.
marine_animals <- c("whale", "shark", "fish", "snail")

# print marine_animals
marine_animals

# Now, let's try a vector of numbers. Let's say that we measured a bunch of different fish and recorded their weights in kilograms.
weight <- c(100, 50, 20, 4, 1, 3, 60)

# Let's say "small" fish are any fish that are < 10 kilograms. We want to know the weights of all the "small" fish that we collected.
weight[weight < 10]

# What if we want all the "big" fish?
weight[weight > 10]
```

Data Manipulation using dplyr and tidyr 
(taken from: https://gge-ucd.github.io/R-DAVIS/lesson_dplyr_ecology.html#Reshaping_with_pivot_longer_and_pivot_wider)

 + Bracket subsetting is handy, but it can be cumbersome and difficult to read, especially for complicated operations. Enter dplyr. dplyr is a package for making tabular data manipulation easier. It pairs nicely with tidyr which enables you to swiftly convert between different data formats for plotting and analysis.

 + Packages in R are basically sets of additional functions that let you do more stuff. The functions we’ve been using so far, come built into R; packages give you access to more of them. Before you use a package for the first time you need to install it on your machine, and then you should import it in every subsequent R session when you need it. 

 + This is an “umbrella-package” that installs several packages useful for data analysis which work together well such as tidyr, dplyr, ggplot2, tibble, etc.

The tidyverse package tries to address 3 major problems with some of base R functions: 
    1. The results from a base R function sometimes depend on the type of data. 
    2. Using R expressions in a non standard way, which can be confusing for new learners. 
    3. Hidden arguments, having default operations that new learners are not aware of.

### 0. Set up

First, we'll load in our packages (libraries) and data. We'll add in a chunk of code to do this.

The shortcut for adding code chunks in RMarkdown is Cmd + Option + I (Mac) or Ctrl + Alt + I (Windows). I've already put the code chunks where they're supposed to go, but practice using this shortcut.

```{r warning = FALSE, message = FALSE}
# you can comment on your code by putting a pound sign at the beginning of the line
# this greys out as an indication that it is not code that will run

# libraries
library(tidyverse)
library(janitor)

# let's get our working directory (where we can directly call our files to import into R)
getwd()
# can also check the file pane (bottom left, usually) and select the file

# data
urchins <- read_csv("sea-urchin.csv")

urchins

# Notice that the class of the data is now tbl_df This is referred to as a “tibble”. Tibbles are almost identical to R’s standard data frames, but they tweak some of the old behaviors of data frames. For our purposes the only differences between data frames and tibbles are that:

# When you print a tibble, R displays the data type of each column under its name; it prints only the first few rows of data and only as many columns as fit on one screen.
```

Today we'll be relying on the [Tidyverse](https://www.tidyverse.org/) for most of our data exploration and visualization. One of the most useful tools in the tidyverse is the pipe operator, which looks like this: `%>%`. This allows you to string together functions, and we'll be using it a lot in this workshop. The shortcut for the operator is Cmd + Shift + M (Mac) and Ctrl + Shift + M (Windows).

### 1. Cleaning up

If you look at the object `urchins`, you'll see that the column names have capitals and spaces. We don't want to have to type all those out, so we can use the function `janitor::clean_names()` to clean them up.

```{r}
urchins2 <- urchins %>% 
  clean_names() 

urchins2

head(urchins)
head(urchins2)
colnames(urchins)
colnames(urchins2)
```

### 2. Basic statistics

The first step to data analysis is basic statistics. These allow you to gather information to calculate statistical relationships down the line. We'll start by calculating the average mass of urchins by species on the reefs.

```{r}
# use urchins2
av_mass <- urchins2 %>% 
  summarize(mean = mean(sea_urchin_mass_g))  # use summarize() to calculate the mean sea urchin mass

av_mass
```

This is not exactly what we want - `av_mass` is now the average mass of all urchins we measured. The equivalent in Excel would be if you had calculated the mean of all the values in the column for mass and had not used the filter tabs to select the values that you wanted.

We can filter values in R (reproducibly!) using `dplyr::filter()`.

```{r}
EM_only <- urchins2 %>% 
  filter(sea_urchin_species == "Echinometra mathaei")

View(EM_only)

EM_mass <- EM_only %>% 
  summarize(mean = mean(sea_urchin_mass_g))

View(EM_mass)

# we can string these together using the pipe operator!
EM_mass2 <- urchins2 %>% 
  filter(sea_urchin_species == "Echinometra mathaei") %>% 
  summarize(mean = mean(sea_urchin_mass_g))

EM_mass2
```

Let's do the same thing for the other urchin species, _Diadema savignyi_. 
```{r}
DS_mass <- urchins2 %>% 
  filter(sea_urchin_species == "Diadema savignyi") %>% 
  summarize(mean = mean(sea_urchin_mass_g))

DS_mass
```

We went through the same steps of filtering for the species, then calculating the mean. This is easy enough with 2 species of urchin, but what if you had 20 different species? You wouldn't want to do this for each of those species individually - it would be way too messy. However, we can use a function called `dplyr::group_by` to tell R, "I want you to consider these groups in my data, and apply a function to these groups separately." 

Let's do that with the urchin data.

```{r}
av_mass_all <- urchins2 %>% 
  group_by(sea_urchin_species) %>% # this doesn't change anything about the data, but stores information about groups in R's brain
  summarize(mean = mean(sea_urchin_mass_g))

av_mass_all
```

We're getting closer. We want to have average masses of each species at each site, but we can use `group_by()` in the same way.

```{r}
av_mass_site <- urchins2 %>% 
  group_by(reef_habitat, sea_urchin_species) %>% 
  summarize(mean = mean(sea_urchin_mass_g))

av_mass_site
```

Now we have means, but what about standard error?

```{r}
final <- urchins2 %>% 
  group_by(reef_habitat, sea_urchin_species) %>% 
  summarize(mean = mean(sea_urchin_mass_g), # first calculate mean
            err = sd(sea_urchin_mass_g)/sqrt(length(sea_urchin_mass_g))) # then calculate standard error

final
```

### 3. Visualization

To create a bar graph like we did in class, we'll use the `ggplot2` package in the Tidyverse. You can make graphs in baseR (the built-in option), but `ggplot2` allows a lot more customization.

```{r}
mass_plot <- ggplot(final, aes(x = reef_habitat, y = mean, fill = sea_urchin_species)) +
  geom_col(position = "dodge", width = 0.8) +
  geom_errorbar(aes(ymin = mean - err, ymax = mean + err), position = position_dodge(0.8), width = 0.3)

mass_plot
```

In this example, the `fill` call tells R, "These are groups that I want to graph separately."

We won't go through all the code for this graph, but you can format this professionally for publication using functions within `ggplot()`.

```{r}
mass_plot_pub <- ggplot(final, aes(x = reef_habitat, y = mean, fill = sea_urchin_species)) +
  geom_col(position = "dodge", width = 0.8, color = "black") +
  geom_errorbar(aes(ymin = mean - err, ymax = mean + err), position = position_dodge(0.8), width = 0.3) +
  scale_y_continuous(expand = c(0,0), limits = c(0, 24)) +
  scale_fill_manual(values = c("#616161", "#adadad")) +
  theme_minimal() +
  labs(x = "Reef Habitat",
       y = "Average mass (g)", 
       fill = "Urchin species")

mass_plot_pub
```

Now, we can create the same scatter plot that we did in class. The first thing we'll have to do is filter the `urchins2` data frame to only include observations from the back reef.

```{r}
back_reef <- urchins2 %>% 
  filter(reef_habitat == "Back Reef")

back_reef
```

Now we can use this data frame to feed into ggplot.

```{r}
scatter <- ggplot(back_reef, aes(x = sea_urchin_mass_g, y = sea_urchin_spine_length_cm)) +
  geom_point() +
  geom_smooth(method = lm)

scatter
```

Here's the code for making this plot look professional:

```{r}
scatter_pub <- ggplot(back_reef, aes(x = sea_urchin_mass_g, y = sea_urchin_spine_length_cm)) +
  geom_point(size = 2) +
  geom_smooth(method = lm, colour = "black") +
  theme_minimal() +
  labs(x = "Sea urchin mass (g)",
       y = "Sea urchin spine length (cm)")

scatter_pub
```

### 4. Cooler visualizations

What if you wanted to see what trends between mass and spine length were for both species in the back reef?

```{r}
scatter_new <- ggplot(urchins2, aes(x = sea_urchin_mass_g, y = sea_urchin_spine_length_cm, group = sea_urchin_species)) +
  facet_wrap(~reef_habitat) +
  geom_point(aes(shape = sea_urchin_species), size = 2, alpha = 0.8) +
  scale_shape_manual(values = c(16, 2)) +
  geom_smooth(method = lm, color = "red") +
  theme_minimal() +
  labs(x = "Sea urchin mass (g)", 
       y = "Sea urchin spine length (cm)",
       shape = "Sea urchin species")

scatter_new
```

https://rstudio.com/wp-content/uploads/2015/03/ggplot2-cheatsheet.pdf

http://sape.inf.usi.ch/quick-reference/ggplot2/colour

Other useful functions in dplyr
```{r}
# select
urchins3 <- select(urchins2, reef_habitat, sea_urchin_species, sea_urchin_mass_g)
urchins3

# mutate
urchins4 <- urchins2 %>%
  mutate(standardized_mass_g = (sea_urchin_mass_g - mean(sea_urchin_mass_g)) / sd(sea_urchin_mass_g))  %>%
  mutate(standardized_spine_cm = (sea_urchin_spine_length_cm - mean(sea_urchin_spine_length_cm) / sd(sea_urchin_spine_length_cm) )) 

urchins4

scatter_new2 <- ggplot(urchins4, aes(x = standardized_mass_g, y = standardized_spine_cm, group = sea_urchin_species)) +
  facet_wrap(~reef_habitat) +
  geom_point(aes(shape = sea_urchin_species, color = sea_urchin_species), size = 2, alpha = 0.8) +
  scale_shape_manual(values = c(21, 22)) +
  geom_smooth(method = lm, color = "red") +
  theme_minimal() +
  labs(x = "Sea urchin mass (g)", 
       y = "Sea urchin spine length (cm)",
       shape = "Sea urchin species", 
       color = "Sea urchin species") 

scatter_new2


scatter_new
```